Effects of the COVID-19 pandemic on economic well-being in the United States

### 2022-03-18 outputs/paper/paper.Rmd
add 2018 data, more writeup and update gitignore

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

@@

Another significant change was that treatment non-respondents was not handled

@@

For the purpose of this paper, we will be focusing on the sub-section of economic well-being within the replicating core, specifically in the 2018 and 2021 US GSS.

# Discussion

## First discussion point

## Second discussion point

## Third discussion point

## Weaknesses and next steps

- difference in survey methodology
- survey format in person vs 
- 2021 codebook stating higher than normal non-response for factual information

>>>>

```{r setup, include=FALSE}
# hold figures in place solution (header_fix.tex) from  https://stackoverflow.com/a/33801326
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "!H")
library(here)
library(tidyverse)
```

@@

Another significant change was that non-respondents was not handled through sub-sampling.

@@

For the purpose of this paper, we will be focusing on the sub-section of economic well-being within the replicating core, specifically in the 2018 and 2021 rounds of the GSS. Of interest are the following questions:

Changes in financial situation - "During the last few years, has your financial situation been getting better, worse, or has it stayed the same?"

Opinion on family income - "Compared with American families in general, would you say your family income is far below average, below average, average, above average, or far above average?"

Standard of living will improve - "The way things are in America, people like me and my family have a good chance of improving our standard of living -- do you agree or disagree?"

Satisfaction with financial situation - "We are interested in how people are getting along financially these days. So far as you and your family are concerned, would you say that you are pretty well satisfied with your present financial situation, more or less satisfied, or not satisfied at all?"

# Discussion
```{r}
gss_data <- read_csv(here::here("inputs/data/gss_data.csv"))
```

## Changes in financial situation

```{r finalter, fig.align="center", fig.height = 6, fig.width = 5}
# get proportion that each response represents, for corresponding age group and year
finalter <- gss_data %>% 
  drop_na() %>% 
  count(age, year, finalter) %>%
  group_by(age, year) %>%
  mutate(pct = (n / sum(n)) * 100)

# set levels of variable, so that it is ordered from best -> worst in facet_wrap
finalter <- finalter %>% 
  mutate(finalter = factor(finalter, levels = c('Getting better', 'Stayed the same', 'Getting worse')))

# plot the percentages, facet_wrap to separate plots by opinion
finalter %>% ggplot(aes(x = year, y = pct, colour = age)) + 
  geom_point() + 
  geom_line() +
  scale_x_discrete(limits=c(2018, 2021)) +
  facet_wrap(~finalter, scales = 'free_y', ncol = 1) +
  labs(title = 'Opinions of financial situation', x = 'Year', y = 'Percentage of group with opinion (%)', color = 'Age group') + 
  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
  theme_linedraw() + 
  scale_colour_brewer(palette = "Set1")
```
- 18-29 age group saw the most significant change in opinion, from being the group who least considered their financial situation to be getting worse to the most.
- All groups had saw a decrease in the response of 'stayed the same', indicating some change
- 40-49, 64-89 saw an increase in 'getting better'; 50-64 almost no change between years (38.69% in 2018, 38.93% in 2021)

## Opinion on family income
```{r finrela}
# get proportion that each response represents, for corresponding age group and year
finrela <- gss_data %>% 
  drop_na() %>% 
  count(age, year, finrela) %>%
  group_by(age, year) %>%
  mutate(pct = (n / sum(n)) * 100)

# set levels of variable, so that it is ordered from best -> worst in facet_wrap
finrela <- finrela %>% 
  mutate(finrela = factor(finrela, levels = c('Far above average', 
                                              'Above average', 
                                              'Average',
                                              'Below average',
                                              'Far below average')))

# plot the percentages, facet_wrap to separate plots by opinion
finrela %>% ggplot(aes(x = year, y = pct, colour = age)) + 
  geom_point() + 
  geom_line() +
  scale_x_discrete(limits=c(2018, 2021)) +
  facet_wrap(~finrela, scales = 'free_y', ncol = 1) +
  labs(title = '', x = 'Year', y = 'Percentage of group with opinion (%)', color = 'Age group') + 
  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
  theme_linedraw() + 
  scale_colour_brewer(palette = "Set1")
```
- note that this is not objective, reliant on subject opinion
- all age groups became more opinionated in 2021- decrease in 'average' response
- larger portion of all groups consider income to be far above average and above average in 2021
- below average- increases for 18-29 and 64-89
- far below average - increases in 18-29, 30-39, 40-49

## Standard of living will improve
```{r goodlife}
# get proportion that each response represents, for corresponding age group and year
goodlife <- gss_data %>% 
  drop_na() %>% 
  count(age, year, goodlife) %>%
  group_by(age, year) %>%
  mutate(pct = (n / sum(n)) * 100)

# set levels of variable, so that it is ordered from best -> worst in facet_wrap
goodlife <- goodlife %>% 
  mutate(goodlife = factor(goodlife, levels = c('Strongly agree', 
                                              'Agree', 
                                              'Neither agree nor disagree',
                                              'Disagree',
                                              'Strongly disagree')))

# plot the percentages, facet_wrap to separate plots by opinion
goodlife %>% ggplot(aes(x = year, y = pct, colour = age)) + 
  geom_point() + 
  geom_line() +
  scale_x_discrete(limits=c(2018, 2021)) +
  facet_wrap(~goodlife, scales = 'free_y', ncol = 1) +
  labs(title = '', x = 'Year', y = 'Percentage of group with opinion (%)', color = 'Age group') + 
  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
  theme_linedraw() + 
  scale_colour_brewer(palette = "Set1")
```

## Satisfaction with financial situation
```{r satfin}
# get proportion that each response represents, for corresponding age group and year
satfin <- gss_data %>% 
  drop_na() %>% 
  count(age, year, satfin) %>%
  group_by(age, year) %>%
  mutate(pct = (n / sum(n)) * 100)

# set levels of variable, so that it is ordered from best -> worst in facet_wrap
satfin <- satfin %>% 
  mutate(satfin = factor(satfin, levels = c('Pretty well satisfied', 'More or less satisfied', 'Not satisfied')))

# plot the percentages, facet_wrap to separate plots by opinion
satfin %>% ggplot(aes(x = year, y = pct, colour = age)) + 
  geom_point() + 
  geom_line() +
  scale_x_discrete(limits=c(2018, 2021)) +
  facet_wrap(~satfin, scales = 'free_y', ncol = 1) +
  labs(title = '', x = 'Year', y = 'Percentage of group with opinion (%)', color = 'Age group') + 
  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
  theme_linedraw() + 
  scale_colour_brewer(palette = "Set1")
```
- more divisive and mixed opinions than prior questions

## Weaknesses

- difference in survey methodology
- survey format in person vs web/phone
- question wording for financial changes: "last few years", recency bias?
- 2021 codebook stating higher than normal non-response for factual information, response bias

## Next steps

- under representation of certain populations in 2021 GSS, using weights to adjust
- missing questions from the replicating core in 2021, specifically regarding job security
